---
title: "01_process_idats"
author: "Puvvula"
date: "2023-06-30"
output: pdf_document
---

```{r}
library(pacman)
pacman::p_load(tidyverse, janitor, stringi, UpSetR)
pheno <- "~/Documents/methylation_pilot/"
idats<- "~/Documents/methylation_pilot/idat_files/"
```

#read chip and de-identified subject info
```{r}
pheno <- read_csv(paste0(pheno, "meth_pheno.csv")) |>
  clean_names() |>
  mutate(gsm_id = paste(sentrix_id, sentrix_position, sep = "_")) 

unique_counts <- pheno |>
  group_by(sample_type) |>
  summarise(unique_participants = n_distinct(participant_id),
            n_chips = n_distinct(sentrix_id))
```

#visualize chip info
```{r}
# Generate the heatmap using ggplot2
cb_palette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7",
                   "#999999", "#A69F00", "#6AB4E9", "#109E73", "#E0E442", "#4072B2", "#B55E00", "#EC79A7")

ggplot(pheno, aes(x = sample_type, y = participant_id, fill = factor(sentrix_id))) +
  geom_tile(width = 0.5, height = 0.8,) +
  scale_x_discrete(expand = c(0, 0))+
  scale_fill_manual(values = cb_palette)+
  labs(x = "Sample Type", y = "Subject ID", fill = "Sentry ID") +
  theme_minimal()+
  guides(fill = "none")
```

#================= Mehtylation data ======================#
```{r}
pacman::p_load(minfi, ENmix, IlluminaHumanMethylation450kmanifest, 
               IlluminaHumanMethylation450kanno.ilmn12.hg19, FlowSorted.Blood.450k, 
               sva, wateRmelon)
```

#read idats
```{r}
#select one type of sample to process
pheno_by_tiss <- pheno |> filter(sample_type == "CBMC")

#read idats
idat_files<-  paste0(idats, pheno_by_tiss$gsm_id)
rgset = read.metharray(basenames=idat_files)

#snp info
snp<- getSnpInfo(rgset, snpAnno = NULL)
#find a way to drop snps

# get the 450k annotation data
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

#p-value set
detP <- detectionP(rgset)
```

#different probe removal method - not using this time
```{r}
# https://doi.org/10.1016/j.envint.2023.107774
#Failed probes based on Illumina’s recommendations were removed (detection p-value > 0.01 in 5 % of individuals or more)
# 4,884 probes
#probes_to_remove <- rowSums(detP > 1e-7) > (0.05*ncol(bmiq_normal))
#table(probes_to_remove)

#dropping 4,884 probes
#bmiq_normalFlt_1 <- bmiq_normal[!probes_to_remove,]
```

#remove probes by p-value threshold
```{r}
# Drop probes if at least one subject with high detection p-value
probes_to_remove <- rowSums(detP > 1e-7) > 0
table(probes_to_remove)

#dropping 13,413 probes based on p-value threshold
rgset_f1 <- rgset[!probes_to_remove, ]

# Drop samples if 5% of probes are above the p-value threshold
samples_to_remove <- colSums(detP > 1e-7) > (0.05 * nrow(rgset))
table(samples_to_remove)

# Remove samples exceeding the threshold from the methylation data
rgset_f2 <- rgset_f1[!samples_to_remove, ]
```

#removing probes at sex chromosomes, SNPs, and cross-hybridizing probes
```{r}
mset <- preprocessNoob(rgset_f2)
GRset <- mapToGenome(mset)

#drop snp
GRset_2 <- dropLociWithSnps(GRset, snps=c("SBE","CpG"), maf=0)

# remove sex chromosomes
keep <- !(rownames(GRset_2) %in% ann450k$Name[ann450k$chr %in% c("chrX","chrY")])
table(keep)

#drop 11,458 probes
GRset_3 <- GRset_2[keep,]


# cross-hybrid probe list from: https://www.tandfonline.com/doi/full/10.4161/epi.23470
cross_hyb_probes <- read_csv(url("https://epigen.ccm.sickkids.ca/sample-report/data/quality_control/cross_reactive_probes.csv"),
                             col_names = c("cg_site"))

GRset_4 <- GRset_3[!(rownames(GRset_3) %in% cross_hyb_probes$X1), ]
```


#Methylset prep: standardization, probe, dye, and batch correction
```{r}
#background correction and dye bias normalization
mset <- preprocessNoob(rgset)
densityPlot(getBeta(mset), sampGroups = pheno$sentrix_id, legend = F, main="preprocessNoob")

#Beta MIxture Quantile dilation (BMIQ; wateRmelon) was applied to correct probe design bias
#beta range: 0.002199932 0.999835795 & median = 0.65
bmiq_normal <- BMIQ(mset)
densityPlot(bmiq_normal, sampGroups = pheno$sentrix_id, legend = F, main="preprocessNoob + BMIQ")

#combat batch correction + BMIQ - over adjusting: beta range -0.3956759  1.4982748
#batch correction alone: beta range = -0.3920262  1.4166083
#batchCorrectedData <- ComBat(dat = getBeta(MSet), batch = pheno_by_tiss$sentrix_id)
#qc plot
#densityPlot(batchCorrectedData, sampGroups = pheno$sentrix_id, legend = F)

#check order of samples
matches <- map_lgl(pheno_by_tiss$gsm_id, ~ .x %in% colnames(bmiq_normal))
```





#generating genomic methyl and genomic ratio set
```{r}
genomic_meth_set<- mapToGenome(MSet)
annotation<-getAnnotation(genomic_meth_set) 
annotation=annotation[intersect(rownames(batchCorrectedData),rownames(annotation)),]
#sex prediction
sex_prediction <- getSex(genomic_meth_set)
sex_pred<- tibble(gsm_id= row.names(sex_prediction), pred_sex= sex_prediction$predictedSex)

genomic_ratio_set<- ratioConvert(genomic_meth_set)

manifest_set <- getManifest(MSet)

SNPs_probes<- getSnpBeta(rgset)
```

#reference based cell count estimation from rgset
```{r}
minfi.LC<- estimateCellCounts(rgset, compositeCellType="Blood")
```
