---
title: "10.1_ewas_for_revision"
author: "Puvvula"
date: "2024-02-06"
output: pdf_document
---

```{r}
library(pacman)
pacman::p_load(tidyverse, janitor, CpGassoc, future, future.apply, doParallel)
#processed<- "~/Documents/methylation_pilot/processed/mp_ref_cell_cnt/"
processed<- "~/Documents/methylation_pilot/processed/"
```

```{r}
dat_chem_qn <- read_csv("~/Documents/methylation_pilot/exp_processed/chem_qn_fin.csv") |>
  select(c("participant_id" , 
           "bpf",
           "b_pb",
           "m_cpp", 
           "m_hi_nch", "mcoch",
           "tcc", "tcs",
           "mom_age", "mom_race", "mom_edu", "pgtob", "bmi", "sex")) |>
  mutate(tcc = if_else(tcc < 0.1, 1, 2),
         tcs = if_else(tcs < 1.7, 1, 2),
         bpf = if_else(bpf < 0.2, 1, 2),
         b_pb = if_else(b_pb < 0.1, 1, 2),
         m_cpp = if_else(m_cpp < 0.4, 1, 2),
         mcoch = if_else(mcoch < 0.2, 1, 2),
         m_hi_nch = if_else(m_hi_nch < 0.4, 1, 2)) |>
  mutate_at(vars(2:8), as.factor)
```


#CMBC analysis
```{r, message=FALSE, warning=FALSE}
#load one of these files: cbmc_processed.rda; fp_processed.rda; mp_processed.rda

load(paste0(processed, "fp_processed.rda"))

dat_pheno <- fp_pheno |>
  left_join(dat_chem_qn, by = "participant_id") |>
  mutate(across(where(is.character), as.factor)) 
```

#check sequence of gsm_id from preprocessed and pheno data
```{r}
bVals <- bVals[, match(dat_pheno$gsm_id, colnames(bVals)), drop = FALSE]
mVals <- mVals[, match(dat_pheno$gsm_id, colnames(mVals)), drop = FALSE]

setdiff(dat_pheno$gsm_id, colnames(bVals))
setdiff(dat_pheno$gsm_id, colnames(mVals))

nCpG = dim(bVals)[1]
```

# beta input dataset transforms into m values set in the function
```{r}
#cell types for CMBC: "CD8T", "CD4T", "NK", "Bcell", "Mono"
#cells for placenta: "cell_mp_Trophoblasts", "cell_mp_Stromal", "cell_mp_Hofbauer", "cell_mp_Endothelial", "cell_mp_nRBC"

ewas_loop_pll(exposures= c("bpf",
                           "b_pb",
                           "m_cpp", 
                           "m_hi_nch", "mcoch",
                           "tcc", "tcs"), 
              covariates= c("mom_age", "mom_race", "mom_edu", "pgtob", "bmi", 
                            "cell_fp_Trophoblasts", "cell_fp_Stromal", "cell_fp_Hofbauer", "cell_fp_Endothelial",
                            "cell_fp_nRBC"),
              bVals_list = list(bVals),
              bVals_names = list("fp"),
              dat_pheno = dat_pheno,
              outputFolder = "~/Documents/methylation_pilot/for_peer_review")
```

#summary
```{r}  
summary_result <- ewas_summary("~/Documents/methylation_pilot/for_peer_review")
```


#for manhattan and volcano plot
```{r}
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19) |> 
  as.data.frame() |>
  select(c("UCSC_RefGene_Name","UCSC_RefGene_Group","Relation_to_Island","chr","pos")) |>
  rownames_to_column(var = "cpg_labels")
```


#add annotation information to select ewas output files
```{r}
annotate_and_save(file_names = c("df_manhat_m_cpp_fp.csv",
                                 "df_manhat_m_cpp_mp.csv"),
                  input_folder = "~/Documents/methylation_pilot/for_peer_review",
                  output_folder = "~/Documents/methylation_pilot/for_peer_review/res_annotated_cell_cnt")
```

#performing gene enrichment analysis using KEGG database
```{r}
perform_gene_enrichment(input_folder = "~/Documents/methylation_pilot/for_peer_review/res_annotated_cell_cnt",
                        output_folder = "~/Documents/methylation_pilot/for_peer_review/res_annotated_cell_cnt")

joe_gene_enrichment(input_folder = "~/Documents/methylation_pilot/for_peer_review/res_annotated_cell_cnt",
                        output_folder = "~/Documents/methylation_pilot/for_peer_review/res_annotated_cell_cnt/joe_enrichment")
```


#####################

# Define a function to read a CSV file and return it as a tibble
```{r}
read_csv_as_tibble <- function(file_path) {
  read_csv(file_path)
}
```

# List all CSV files in a directory and combine them into a tibble
```{r}
combined_data <- list.files(path = "~/Documents/methylation_pilot/for_peer_review/res_annotated_cell_cnt/annotated", 
                            pattern = "\\.csv$", full.names = TRUE) |>
  map_dfr(~ {
    read_csv_as_tibble(.) |>
      mutate(
        filename = gsub("^.*/|\\.csv$|^anno_df_manhat_", "", basename(.))
      )
  })
```

#count frequency of cpgs and genes for text
```{r}
summary_df <- df |>
  filter(fdr<0.05,
         str_detect(UCSC_RefGene_Group, "TSS")) |>
  group_by(tissue,UCSC_RefGene_Name, UCSC_RefGene_Group) |>
  summarise(
    n = n(),
    distinct_cpg_labels = toString(unique(cpg_labels)),
    num_unique_cpg_labels = length(unique(cpg_labels)),
    distinct_chemicals = toString(unique(chemical))
  ) |>
  ungroup() |>
  filter(UCSC_RefGene_Name != "NA")
```

#export statistically significant ANOVA results (raw p-value<0.05 not fdr) - for supplement
```{r}
anova_df<- combined_data |>
  filter(p_value<0.05) |>
  mutate(methylation_source = case_when(
    str_sub(filename, nchar(filename) - 1) == "fp" ~ "fetal placenta",
    str_sub(filename, nchar(filename) - 1) == "mp" ~ "maternal placenta",
    TRUE ~ "other"
  )) |>
  mutate(chemical_biomarker = "Mono carboxyisononyl phthalate") |>
  select(-c(filename))

write_csv(anova_df, "~/Documents/methylation_pilot/for_peer_review/export_ewas_results/anova.csv")
```

#export statistically significant (raw p-value < 0.05 not the fdr) linear regression results - for supplement
```{r}
lm_res<- df |>
  mutate(chemical_biomarker = recode(chemical,
                            "bpa" = "Bisphenol-A",
                            "fluo2" = "2-Hydroxyfluorene",
                            "m_cnp" = "Mono carboxyisononyl phthalate",
                            "nap1" = "1-Hydroxynaphthalene",
                            "phen4" = "4-Hydroxyphenanthrene",
                            "pyr1" = "1-Hydroxypyrene",
                            "e_pb" = "Ethyl paraben"))|>
  mutate(methylation_source = recode(tissue, 
                                     "cmbc" = "Cord blood mono nuclear cells",
                                     "fp" = "fetal placenta",
                                     "mp" = "maternal placenta")) |>
  select(-c(filename, chemical, tissue))

write_csv(lm_res, "~/Documents/methylation_pilot/for_peer_review/export_ewas_results/lm_res.csv")
```

